---
title: "Post-Impute"
author: "Mari Johnson"
date: "12/10/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Updates October: - Re-indexed and merged bcf files to get a more
stringent R2 - Making separate typhoid, para and merged plink files

#### Set Up

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
knitr::opts_knit$set(root.dir = '~/GWAS_22/gwas_final/meta')
```

```{r Set up, eval=TRUE}
library(data.table)
library(stringr)
library(dplyr)
library(plinkQC)
library(tidylog)
library(janitor)


load(file = "id_covars.RData")
```

### Check Merged data fam IDs

```{r Tidy-fam-IID, eval = FALSE}
fam <- fread("merged_data.fam")
# Clean IID col to match ID in covar files
fam2 <- fam
fam2$V2 <-  str_remove_all(fam$V2, "^([0-9]+_)")
update_ids <- bind_cols(fam$V1, fam$V2, fam2$V1, fam2$V2)

write.table(update_ids, "update_ids.txt", col.names = F, row.names = F, quote = F, sep = '\t')
system("plink --bfile ~/GWAS_22/gwas_final/merge/merged_data --update-ids ~/GWAS_22/gwas_final/merge/update_ids.txt --make-bed --out ~/GWAS_22/gwas_final/merge/merge_rn")
#TODO let plink2 know about error
```

### Update covar and pheno files

**Key** Strain: C, P = 3, T = 1, TN = 2 Sex M=1, F=2 Re-challenge N=1,
PP=2, PT=3, TP=4, TT=5 Vaccine: NONE = 0, Ty21a = 1, M01ZH09= 2, Vi-PS =
3, Vi-TT= 4

```{r typhoid}
'%!in%' <- function(x,y)!('%in%'(x,y))
# covar checks -----------------------------------------------------------------
# New fam file -re filtered R2
fam <- fread("~/GWAS_22/gwas_final/merge/merge_rn.fam")
covar <- fread("~/GWAS_22/new_gwas/meta/final_metadata/all_covar_june.txt")
# Tidy covar table
covar$chall_vax <- str_c(covar$Challenge, covar$Vaccine, sep = "_")
covar <- clean_names(covar)
covar <- covar %>% dplyr::rename(IID = iid)

# Filter for serovars
covar_typhoid <- covar %>% filter(challenge != 3, re_challenge  != 5) #259
covar_para <- covar %>% filter(challenge == 3, re_challenge  != 2) #71
covar_enteric <- covar %>% filter(re_challenge == 1)  # 312
```

#### Missing data

```{r check-missing}
# Check for missing samples ----------------------------------------------------
fam <- fam %>% rename(IID = V2)
new_covar <- left_join(fam, covar, by = "IID")
# missing samples
missing <- filter(new_covar, is.na(study))
# Make new IID col to match old IIDs 'Genotyping ID'
missing$IID_2 <- str_remove_all(missing$IID, "^([0-9]+-)")
```

```{r Tidy-missing}
 # Tidy old covar files---------------------------------------------------------
check <- fread("~/GWAS_22/new_gwas/meta/covar.txt")
#pheno2 <- fread("~/GWAS_22/new_gwas/meta/pheno.txt")
#covar2 <- left_join(pheno2, covar2, by ="GENOTYPING_ID")
#rm(pheno2)
check <- rename(check, IID = GENOTYPING_ID)
# Look in old covar for IIDs that are in our missing IIDs
check <- filter(check, IID %in% missing$IID_2)
check <-  distinct(check) %>% filter(!is.na(Study))
# Found 5 samples :) with diagnosis info

# Tidy colnames/vars so can rejoin----------------------------------------------
check$Vaccine <- ifelse(check$Vaccine == "None",0,
                            ifelse(check$Vaccine == "Ty21a",1,
                                   ifelse(check$Vaccine == "M01ZH09", 2,
                                          ifelse(check$Vaccine == "Vi-PS", 3,
                                                 ifelse(check$Vaccine == "Vi-TT", 4,NA)))))
#Recode strain, sex challenge
check$Challenge <- ifelse(check$Challenge == "P", 3, ifelse(check$Challenge == "T", 1, NA))
check$Sex <- ifelse(check$Sex == "M", 1, ifelse(check$Sex == "F", 2, NA))
check$Re_Challenge <- as.numeric(str_replace_all(check$Re_Challenge, "N", "1"))
check$Chall_Vax <- str_c(covar$Challenge, covar$Vaccine, sep = "_")
check <- clean_names(check)
check <- rename(check, IID_2 = iid) # will match IID_2 in missing then tidy
```

```{r Rejoin and remake tables}
# Rejoin -----------------------------------------------------------------------
missing <- missing %>% select(IID, IID_2)
missing <- left_join(missing, check, by = "IID_2") %>% filter(!is.na(study)) %>% select(-IID_2)

#add back to covar
covar <- bind_rows(covar, missing)
#check samples again
test <- anti_join(fam, covar, by = "IID") # 4 missing covars so improved :)

covar_typhoid <- covar %>% filter(challenge != 3, re_challenge  != 5) #263
covar_para <- covar %>% filter(challenge == 3, re_challenge  != 2) #72
covar_enteric <- covar %>% filter(re_challenge == 1)  #317
```

#### Phenotype data

```{r Get-phenodata}
pheno <- fread("~/GWAS_22/new_gwas/meta/final_metadata/all_pheno_num.txt")
pheno <-  filter(pheno, IID %in% covar$IID) # 342

covar_typhoid <- left_join(covar_typhoidpheno, by="IID")
covar_para <- left_join(covar_para,pheno, by="IID")
covar_enteric <- left_join(covar_enteric,pheno, by="IID")
```

#### PATCH samples, and missing sex

```{r Extra PATCH samples}
# Check what fam IDs are in covar_typhoid
test <- filter(fam, IID %in% covar_typhoid$IID)# 255 #redo = 260 :)
# Check what covar_typhoid IIDs are not in test IIDs
patch <- filter(covar_typhoid, IID %!in% test$IID) 
# test <- filter(patch, IID %in% fam$IID) # 5 extra samples
# just need to remove _PATCH string from covar file so it will match fam IID

# Missex samples ---------------------------------------------------------------

missex <- fread("fail_sexcheck.txt")
missex <- missex[,V2]
covar_typhoid <- covar_typhoid %>% filter(IID %!in% missex)

# Make new tables
covar_typhoid$IID <- str_remove_all(covar_typhoid$IID, "_PATCH")

# Get FID back
fam <-  fam %>% select(V1, IID) %>% rename(FID = V1) #254
covar_typhoid <- inner_join(fam, covar_typhoid, by = "IID")
# Filter duplicates
covar_typhoid <- filter(covar_typhoid, stri_duplicated(IID)==FALSE) %>%
  select(-ethnicity)

# Recode vars (plink2 can use categorical)
covar_typhoid$vaccine <- ifelse(covar_typhoid$vaccine == 0,"None",
                            ifelse(covar_typhoid$vaccine == 1,"Ty21a",
                                   ifelse(covar_typhoid$vaccine == 2, "M01ZH09",
                                          ifelse(covar_typhoid$vaccine ==  3,"Vi-PS",
                                                 ifelse(covar_typhoid$vaccine == 4,"Vi-TT",NA)))))


#C, P = 3, T = 1, TN = 2
covar_typhoid$challenge <- as.factor(covar_typhoid$challenge)
levels(covar_typhoid$challenge) <- c("T", "TN")

covar_typhoid$re_challenge  <- as.factor(covar_typhoid$re_challenge)
levels(covar_typhoid$re_challenge) <- c("N", "PT")

covar_typhoid$chall_vax <- str_c(covar_typhoid$challenge, covar_typhoid$vaccine, sep = "x")

# Missing pheno data -----------------------------------------------------------
missing_p <- filter(covar_typhoid, is.na(Diagnosed))
p_data <- fread("pheno_fix.txt") #added zeros back in to IIDs where needed
p_data <- filter(p_data, IID %in% missing_p$IID)
rm(missing_p)

covar_typhoid <- covar_typhoid %>% left_join(p_data, by= "IID") %>%
  mutate(Diagnosed = coalesce(Diagnosed.x, Diagnosed.y))
covar_typhoid <- select(covar_typhoid, -c(Diagnosed.x, Diagnosed.y, place_of_birth))

# Make pheno and keep files ----------------------------------------------------
pheno_typhoid <- covar_typhoid %>% select(FID, IID, Diagnosed)
keep_t <- select(covar_typhoid, FID, IID)

write.table(keep_t, file="keep_t.txt", col.names = T, row.names = F, quote = F, sep = '\t')
write.table(covar_typhoid, file="covar_typhoid.txt",col.names = T, row.names = F, quote = F, sep = '\t')
write.table(pheno_typhoid, file="pheno_typhoid.txt", col.names = T, row.names = F, quote = F, sep = '\t')

#TODO paratyphoid files and enteric

```

#### All enteric samples

```{r enteric}
# Enteric -------------------------------------------------------------------------

fam_ent <- fread("~/GWAS_22/gwas_final//merge/merge_rn.fam")
fam_ent <-  fam_ent %>% select(V1, V2) %>% rename(FID = V1, IID = V2) #254
covar_enteric <- covar_enteric %>% filter(IID %!in% missex)
covar_enteric$IID <- str_pad(covar_enteric$IID, 3, pad = "0")

covar_enteric <- left_join(covar_enteric, fam_ent, by = "IID")


#missing 5 add leading zeros


# Filter duplicates
covar_enteric <- filter(covar_enteric, stri_duplicated(IID)==FALSE) %>%
  select(-ethnicity, -place_of_birth, - re_challenge)

# Recode vars (plink2 can use categorical)
covar_enteric$vaccine <- ifelse(covar_enteric$vaccine == 0,"None",
                            ifelse(covar_enteric$vaccine == 1,"Ty21a",
                                   ifelse(covar_enteric$vaccine == 2, "M01ZH09",
                                          ifelse(covar_enteric$vaccine ==  3,"Vi-PS",
                                                 ifelse(covar_enteric$vaccine == 4,"Vi-TT",NA)))))


#C, P = 3, T = 1, TN = 2
covar_enteric$challenge <- as.factor(covar_enteric$challenge)
levels(covar_enteric$challenge) <- c("T", "TN","P")

covar_enteric$chall_vax <- str_c(covar_enteric$challenge, covar_enteric$vaccine, sep = "x")

# Missing pheno data -----------------------------------------------------------
missing_p <- filter(covar_enteric, is.na(Diagnosed))
p_data <- fread("pheno_fix.txt") #added zeros back in to IIDs where needed

p_data <- filter(p_data, IID %in% missing_p$IID)
rm(missing_p)

covar_enteric <- covar_enteric %>% left_join(p_data, by= "IID") %>%
  mutate(Diagnosed = coalesce(Diagnosed.x, Diagnosed.y))
covar_enteric <- select(covar_enteric, -c(Diagnosed.x, Diagnosed.y))

# Make pheno and keep files ----------------------------------------------------
pheno_enteric <- covar_enteric %>% select(FID, IID, Diagnosed)
keep_ent <- select(covar_enteric, FID, IID)

write.table(keep_ent, file="~/GWAS_22/gwas_final/merge/enteric/keep_ent.txt", row.names = F, quote = F, sep = '\t')
write.table(covar_enteric, file="covar_enteric.txt", row.names = F, quote = F, sep = '\t')
write.table(pheno_enteric, file="pheno_enteric.txt",  row.names = F, quote = F, sep = '\t')

```

#### PCA

```{r pca}
fam <- fread("~/GWAS_22/gwas_final/merge/typhoid/QC/typhoid2.IBD.fam") # 251
pca <- fread("~/GWAS_22/gwas_final/merge/typhoid/QC/pca/typhoid2.LD.eigenvec") # 251
test <- pca %>% filter(IID %in% covar_typhoid$IID) #250
# remake typhoid files
save.image(file = "id_covars.RData")
load(file = "id_covars.RData")
```

```{bash filter-mergern}
plink --bfile /home/mari/GWAS_22/gwas_final/merge/merge_rn \
--keep /home/mari/GWAS_22/gwas_final/meta/keep_t.txt \
```


```{bash filter-mergern}
--make-bed \
--out /home/mari/GWAS_22/gwas_final/merge/typhoid/typhoid2
# 253 participants
```

#### VAST data

```{r vast covar_pheno file}
setwd("~/GWAS_22/gwas_final/merge/typhoid/vast")

# Vast vaccination samples
fam2 <- fread("vast.fam")
pca2 <- fread("vast.LD.eigenvec")

fam2 <- fam2 %>% dplyr::rename(FID = V1, IID = V2) %>% dplyr::select(FID, IID)

vast_covar <- fam2 %>% left_join(covar_typhoid, by = c("FID","IID")) %>% left_join(pca2, by = c("FID","IID"))

vast_pheno <- fam2 %>% left_join(pheno_typhoid, by = c("FID","IID"))

# Merge with ab data ---------------------------------------------------------------
# vast_assoc_notes.R
load(file = "~/GWAS_22/gwas_final/meta/vast-phenos.RData") 
vast_pheno <- left_join(vast_pheno, pheno, by = c("FID","IID"))

write.table(vast_covar, file="covar_vast.txt",row.names = F, quote = F, sep = '\t')
write.table(vast_pheno, file="pheno_vast.txt", row.names = F, quote = F, sep = '\t')
```

#### Update vast vax pheno

```{r vast pheno checking}

check <- clean_names(fread("enteric_meta_Feb22.csv"))
check <- filter(check, !is.na(genotyping_id))

pheno_exprs <- read.csv(file = "~/RNA/oct/combat_vast_tyger.csv")
pheno <- pheno_exprs[, !str_detect(colnames(pheno_exprs), "ENSG")]
pheno <- filter(pheno, time_point == "Baseline")

check2 <- left_join(pheno, check, by = c("lab_id", "participant_id"))
check2 <- dplyr::select(check2, participant_id, genotyping_id, lab_id, meta_id, diagnosis, outcome, study, study_arm, vaccine, strain_or_dose, age_years, sex, ancestry)
# check 2 has pheno data from rnaseq, and old covar data
# check 3, add in current covar data

# VAST ####
vast_check <- filter(check2, study == "VAST")
IDlink <- clean_names(read.csv(file = "geno_ids.csv"))
#Rename ID col so can join appropriately
IDlink <- dplyr::rename(IDlink, lab_id = lab_id)
vast_check <- left_join(vast_check, IDlink, by = c("genotyping_id", "lab_id",  "participant_id", "meta_id"))

# Vaccine mismatches in covar file
covar_typhoid2 <- dplyr::rename(covar_typhoid, cat_iid = IID)

vast_check <- inner_join(covar_typhoid2, vast_check, by = "cat_iid")

# recode vaccine.x (my covar info to match seq info)
vax <- vast_check %>% mutate(vaccine.x = ifelse(vaccine.x == "None", "Control", vaccine.x),
                      vaccine.x = ifelse(vaccine.x == "Vi-TT", "Vi-TCV", vaccine.x))

miss <- vax[vax$vaccine.x != vax$vaccine.y, ] # 16 :(

# Get cols to update covar_typhoid with 
miss <- miss[, c(1:7,27,9:18)]
# tidy_colnames and vars
names(miss) <- str_remove_all(colnames(miss), "\\.\\w")
miss <- miss %>% mutate(vaccine = ifelse(vaccine == "Control", "None", vaccine),
                      vaccine = ifelse(vaccine == "Vi-TCV", "Vi-TT", vaccine))

miss$chall_vax <- str_c(miss$challenge, miss$vaccine, sep = "x")

miss <- rename(miss, IID = cat_iid)


covar_typhoid <- filter(covar_typhoid, IID %!in% miss$IID)
covar_typhoid <- bind_rows(miss, covar_typhoid)
write.table(covar_typhoid, file="covar_typhoid.txt", row.names = F, quote = F, sep = '\t')

# Add pca
pca <- fread("~/GWAS_22/gwas_final/merge/typhoid/QC/pca/typhoid2.LD.eigenvec") # 251
covar_pca <- left_join(covar_typhoid, pca, by = "IID")#250
write.table(covar_pca, file="typhoid_pcacovar.txt", row.names = F, quote = F, sep = '\t')

save.image(file = "id_covars.RData")

# Tyger ####
# Tyger all matches up
# check against current covar
# covar_typhoid2 <- dplyr::select(covar_typhoid, FID, IID, study, age, sex, challenge, vaccine, chall_vax)
# covar_typhoid2 <- dplyr::rename(covar_typhoid2, genotyping_id = IID)
# check3 <- inner_join(check2, covar_typhoid, by = "genotyping_id") 
```

