---
title: "Genetic Associations of Susceptibility to Typhoid Fever"
author: "Mari Johnson"
date: "2022-10-12"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = '~/GWAS_22/gwas_final/merge/assoc')

library(data.table)
library(ggplot2)
library(ggpubr)
library(qqman)
library(tidylog)
library(stringr)

```

### Aims:

-   Using imputed and cleaned data, associate genome-wide SNP frequency in our cohort with susceptibility/protection from typhoid fever following oral challenge (GWAS)
-   Association analysis can be modified to also look at regions of interest (CGAS)
-   Check Q-Q plot and genomic inflation values
-   Visualise GWAS significance with Manhattan plots
-   Interaction testing of covariates
-   Filter and format top hit data for downstream annotation

### Linear Regression GWAS models

Plink2 fits generalised linear mixed effects model:

y = G𝛽 + C𝛽 + e

-   Where, y is our response variable - ReSVinet score
-   G is our genotype dosage matrix
-   C is our covariate matrix
-   e is the error standard error (or residual) of our model estimate
-   𝛽 are our model coefficients of determination. So, for each SNP we are modelling whether zero, one or two copies of an allele in our population, against our linear y variable, RSV severity.

We then test to see if there is a relationship between allele dosage and outcome, how strong the effect is, if it is significant

The null hypothesis is that there is no linear relationship. Therefore if;\
𝛃 = 0 the allele shows no relationship with our phenotype score\
𝛃 \> 0 the allele is associated with a more positive score\
𝛃 \< 0 the allele is associated with a more negative score

The significance of this relationship is tested by calculating the T statistic:\
\
**t = b / SE** - **b** is the estimated value of linear slope or the coefficient of the predictor variable.\
- **SE~*b*~** represents the standard error of the coefficient estimation which can be estimated using the following formula:\
- **SE = S / √N** Where S represents the standard deviation and N represents the total number of data points

If the p-value that corresponds to t is less than some threshold (e.g. α = .05); then we reject the null hypothesis and conclude that there is a statistically significant relationship between the predictor variable and the response variable

### Case-control analysis

-   General linear association model (GLM): Diagnosis
-   Covariates to include: PCs, Sex, <https://www.cog-genomics.org/plink/2.0/assoc>

```{r meta, Check covariates}

```

```{bash GWAS Enteric fever}
plink2 --bfile typhoid.IBD \
--pheno ~/GWAS_22/gwas_final/meta/pheno_typhoid.txt \
--pheno-name diagnosed \
--covar ~/GWAS_22/gwas_final/meta/covar_pca.txt \
--covar-name sex, chall_vax, age, PC1, PC2, PC3, PC4, PC5 \
--covar-variance-standardize \
--glm \
--adjust \
--out T_assoc
```

New PCs lamda (check using --adjust flag) = 1, no genomic inflation :)

#### Interaction testing

Warning message: \--glm remaining control count is less than 10x predictor count for phenotype 'Diagnosed'\
As I now have a much smaller sample size and more covariates I have 10 covariates which will be less than 1/10th of our n 153.

I am already thinking there is probably issues with linear dependencies based on the covar, i.e with dose (poss concat this with challenge).\
To many covars and too small an n can overfit model.\
Therefore test for significance of these factors in linear model.\
Remove least significant covars.

-   covar standardise necessary, but does pass VIF test (=2)

```{bash GWAS Covar checking}
mkdir covar_checks 
plink2 --bfile typhoid.IBD \
--pheno ~/GWAS_22/new_gwas/meta/final_metadata/T_pheno_Jun.txt \
--pheno-name Diagnosed \
--covar ~/GWAS_22/new_gwas/meta/final_metadata/T_covar_Jun.txt \
--covar-name Sex, Challenge, Dose, Vaccine, Age, PC1, PC2, PC3, PC4, PC5 \
--glm genotypic interaction \
--covar-variance-standardize \
--out covar_checks/check

```

```{r Interaction test}

interact <- fread("covar_checks/check.Diagnosed.glm.logistic.hybrid")
colnames(interact) <-
  c("CHR","BP", "SNP","REF","ALT","A1",
    "FIRTH?", "TEST","OBS_CT","OR","LOG(OR)_SE",
    "Z_STAT", "P","ERRCODE")

# Remove unnecessary columns
interact <- interact %>% 
  select(!c(ALT, OBS_CT,`FIRTH?`,`LOG(OR)_SE`, Z_STAT, OR)) 

interact$TEST <- as.factor(interact$TEST)
levels(interact$TEST)

VIF_high <- filter(interact, ERRCODE == "VIF_TOO_HIGH")
table(VIF_high$TEST)

VIF_ADD <- filter(interact, TEST == "ADD")

# 1947595 - VIF too high, are these for low sig snps?
# Filter for top SNPS? Then look at interaction factor sig
# Combine vaccine and challenge into one factor

# Just vaccine 
vax <- interact[(interact$TEST == "Vaccine"),]
vax %>%
  group_by(ID) %>%
  ggplot(aes(
    x = TEST,
    y = -log10(P),
    fill = TEST
  )) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "\nCovariate", title = "Interaction Tests") +
  theme(legend.position = "none")

# ADD
geno <- interact %>% filter(TEST == "ADD")
geno <- filter(geno, P <1e-2) #topsnps = <10e-4
snps <- geno$SNP
top <- filter(ADD, SNP %in% snps)

# glm is fitting a model per snp, therefore get an overview of covars
tiff(file = "interaction_snps.tiff",
     res = 720,
     width = 7500, height = 5000,
     compression = "lzw")
top %>% group_by(SNP) %>%
  ggplot(aes(
    x = TEST,
    y = -log10(P),
    fill = TEST
  )) + geom_boxplot() +
  theme_bw() + labs(x = "\nCovariate", title = "Interaction Tests") + theme(legend.position = "none")
dev.off()


# Make new factor - updated in pca_plot script too
covar$chall_vax <- str_c(covar$Challenge, covar$Vaccine, sep = "_")
covar$chall_vax <- as.factor(covar$chall_vax)
```

### Assoc Results

```{r Plot Assoc Results, eval = TRUE, cache=TRUE}


# Typhod Assoc -----------------------------------------------------------------
gwas <- fread("T_assoc.Diagnosed.glm.logistic.hybrid")

gwas <-
  filter(gwas, TEST == "ADD") 

# Rename cols for qqman
colnames(gwas) <-
  c("CHR","BP", "SNP","REF","ALT","A1",
    "FIRTH?", "TEST","OBS_CT","OR","LOG(OR)_SE",
    "Z_STAT", "P","ERRCODE"
  )

gwas <- select(gwas, !c(`FIRTH?`,ERRCODE)) 

write.table(gwas, file = "gwas_typhoid.txt", quote = F, row.names= F, sep = "\t")


# Inspect top snps: ------------------------------------------------------------

tophits <- filter(gwas, P < 10e-5) 
write.table(tophits, file = "tophits_typhoid.txt", quote = F, row.names= F, sep = "\t")


#### Manhattan Plots

# Plot Manhattan
manhattan(gwas, ylim = c(0, 8), cex = 0.6, cex.axis = 0.9, genomewideline = F, col = c("mediumseagreen", "darkolivegreen4"))

pdf(file =(paste(plot_dir,"manhattan_typhoid.pdf",sep ="")), paper = "a4r")
manhattan(gwas, ylim = c(0, 8), cex = 0.6, cex.axis = 0.9, genomewideline = F, col = c("mediumseagreen", "darkolivegreen4"))
dev.off()

#qq_plot

qq(gwas$P)
# More deflated :(

# See plotting2.R for more data viz
```

```{bash}
# Prep for annotation
# cut in bash

cut -f 3 tophits22.txt > explore_hits.txt
sed -i 's/chr//g' explore_hits.txt
cut -d ':' -f 1,2 explore_hits.txt > explore_hits2.txt

# Updates - server did not work, try fuma
```

```{r gtex}

gtex <- read.csv(file = "gtex.csv")
gtex <- drop_na(gtex) %>% filter(dbSNP != "none")
gtex$tissue <- rep("Lung", length(gtex$dbSNP))
gtex <- gtex[1:400,]
write.csv(gtex, file = "gtex_lung.csv", row.names = F, quote = F)

```
