---
title: "Pre-Imputation Steps"
author: "Mari Johnson"
date: "07/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(engine.opts='-l')
```

## Introduction 

This script includes preparation steps for QC and formatting GWAS microarray data for imputation\
There are 3 eneteric GWAS experiments (across 6 studies) that need to be QC'd and merged;\
- T1 T2
- P1 and Tyger
- VAST and PATCH

Raw data was saved in PLINK1 format, updated to binary format (fam, bed, bim)

VAST_PATCH files have updated IDs which were saved as new bed files\
See ID_updating/New_IDs.R for details\
File Tracker.txt also has details of file types\

## Aims 

- To ensure all studies have the same RS IDs and genome version
- QC data for duplicates, A > T SNPs etc.
- Merge QC'd Study Data
- Format data into vcf.gz files for uploading to Topmed Imputation Server

## Set Up 

```{r Library}
setwd("~/GWAS_22/new_gwas")
library(data.table)
library(stringr)
library(dplyr)
library(tidylog)
library(stringi)
```
```{bash Set up folders}
cd ~/GWAS_22/new_gwas
#still need to troubleshoot setting bashrc profile to run in code chunks engine opts - l didn't work,
#Have moved plink2 to ~/bin which is in the markdown default shell PATH
```

```{bash Make folders}
mkdir ~/GWAS_22/new_gwas/updated_snps ;
mkdir ~/GWAS_22/new_gwas/updated_snps/VAST ;
mkdir ~/GWAS_22/new_gwas/updated_snps/T1 ;
mkdir ~/GWAS_22/new_gwas/updated_snps/Tyger
```

## SNP ID formatting

```{r Update SNP conversion file}
#Update Global Screening array plink file with Snp coordinates using update name function 
remap <- fread("~/GWAS_22/new_gwas/GSA-24v3-0_rsids.txt", header = F)

remap <- remap[, c(2,1)]

#make new dup check column
remap_2 <- remap %>% mutate(dup_check = stri_duplicated(remap$V2))
#rm duplicates
remap_2 <- remap_2 %>% filter(dup_check == FALSE) #removes 9389 rows
#tidy
remap_2 <- remap_2 %>% select(-dup_check)

write.table(file="~/GWAS_22/new_gwas/GSA-24v3-0_rsids_v2.txt", remap_2, sep = "\t", quote=F, row.names = F, col.names = F)
```

```{r Clean experiment SNPs}
setwd("~/GWAS_22/new_gwas/VAST_PATCH")

#read data library(data.table)
bim<-fread("VAST_PATCH_2.bim")

#Filter SNPs in autosomal chromosomes")
chrom <- c(seq(1,23),"X")
bim <- bim[bim$V1%in%chrom,]

#Remove duplicates 
bim.dups<-bim[duplicated(paste(bim$V1, bim$V4)),]
#subset all SNP IDs that are in col one and col 4 
duplicated_snps<-bim.dups$V2

#remove snps where A > T, C > G as PLINK can't tell what strand to merge/read these SNPs from
atcg_snps <- bim$V2[((bim$V5 == "A") & (bim$V6 == "T")) |
                    ((bim$V5 == "T") & (bim$V6 == "A")) |
                    ((bim$V5 == "C") & (bim$V6 == "G")) |
                    ((bim$V5 == "G") & (bim$V6 == "C"))]

#make exclusion SNP list ##
exclude <- c(duplicated_snps, atcg_snps)

write.table(file="exclude_snps_rm.txt", exclude, sep = "\t", quote=F, row.names = F, col.names = F)

system("plink2 --bfile VAST_PATCH_2 --exclude exclude_snps_rm.txt --make-bed --out VAST_rename_rm")
```


```{r Update GSA SNP names in R}
setwd("~/GWAS_22/new_gwas/VAST_PATCH")
#left join remap_2 to bim file merge add change colnames 
bim_rename<-fread("VAST_rename_rm.bim")
colnames(remap_2) <- c("newID", "oldID")
colnames(bim_rename) <- c("chr", "oldID", "V3", "BP", "A1", "A2")

updated <- left_join(remap_2, bim_rename, by = "oldID")
#reorder cols 
updated <- select(updated, chr, newID, V3, BP, A1, A2)
#remove non matched snps
updated <- filter(updated, chr != "NA")
#sort by chromosome and bp
updated <- updated[order(chr, BP),]

write.table(file="VAST_rename.bim", updated, sep = "\t", quote=F, row.names = F, col.names = F)

```
## Liftover
```{bash Make BED file}
cd ~/GWAS_22/new_gwas/VAST_PATCH
#Make BED file:
#Different to Plink .bed format
awk '{print "chr" $1, $4 -1, $4, $2 }' VAST_rename.bim > VAST_rename_rm_lift

```

```{bash Liftover Online}
#Liftover go online to https://genome.ucsc.edu/cgi-bin/hgLiftOver
#submit file at the bottom of the page

#output
#Successfully converted 625337 records
#Conversion failed on 2003 records.
#Download as VAST_38
```

```{bash Updating Liftover SNPs}


cd ~/GWAS_22/new_gwas/Liftover_Results

# ectract mapped variants snp IDs
awk '{print $4}' VAST_38.bed > VAST_38_rm.snps
# ectract updated positions
awk '{print $4, $3}' VAST_38.bed > VAST_38_rm.pos
#update plink file with SNPs that we could extract, and update the snp coordinates using update map function
plink2 --bfile ~/GWAS_22/new_gwas/VAST_PATCH/VAST_rename --rm-dup --make-bed --out VAST_rename_rm_dup

#Error: VAST_rename.bim has a split chromosome. Use --make-pgen + --sort-vars to remedy this.

#exclude duplicates again from rm-dup list
plink2 --bfile VAST_rename_rm --exclude VAST_rename_rm_dup.rmdup.mismatch --make-bed --out VAST_rename_rm_dup
plink2 --bfile VAST_rename_rm_dup --extract VAST_38_rm.snps --update-map VAST_38_rm.pos --make-bed --out VAST_remap

```





```