---
title: "Post-Impute"
author: "Mari Johnson"
date: "11/03/2022"
output: html_document
---

 Updates October:
 - Re-indexed and merged bcf files to get a more stringent R2
 - Making separate typhoid, para and merged plink files


```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
knitr::opts_knit$set(root.dir = '~/GWAS_22/gwas_final/merge')
#Assoc folder switched to after merging
```

```{r Set up, eval=TRUE}
library(data.table)
library(stringr)
library(dplyr)
library(plinkQC)
library(tidylog)
library(janitor)
```

### Check Merge

```{r Tidy-fam-IID}

fam <- fread("merged_data.fam")
# Clean IID col
fam2 <- fam
fam2$V2 <-  str_remove_all(fam$V2, "^([0-9]+_)")

update_ids <- bind_cols(fam$V1, fam$V2, fam2$V1, fam2$V2)

write.table(update_ids, "update_ids.txt", col.names = F, row.names = F, quote = F, sep = '\t')

system("plink --bfile merged_data --update-ids update_ids.txt --make-bed --out merge_rn")
#TODO let plink2 know about error
```

#### Update covar and pheno files

**Key**
Strain: C, P = 3, T = 1, TN = 2
Sex M=1, F=2
Re-challenge N=1, PP=2, PT=3, TP=4, TT=5
Vaccine: NONE = 0, Ty21a = 1, M01ZH09= 2, Vi-PS = 3, Vi-TT= 4

```{r typhoid}
# covar checks
pheno <- fread("~/GWAS_22/new_gwas/meta/final_metadata/all_pheno_num.txt")
covar <- fread("~/GWAS_22/new_gwas/meta/final_metadata/all_covar_june.txt")

#old files
covar2 <- fread("~/GWAS_22/new_gwas/meta/covar.txt")
pheno2 <- fread("~/GWAS_22/new_gwas/meta/pheno.txt")
covar2 <- left_join(pheno2, covar2, by ="GENOTYPING_ID")

# Make new chall_vax variable
covar$chall_vax <- str_c(covar$Challenge, covar$Vaccine)

covar <- clean_names(covar)
covar <- covar %>% dplyr::rename(IID = iid)

# Filter
covar_enteric <- covar %>% filter(re_challenge == 1)
covar_typhoid <- covar %>% filter(challenge != 3, re_challenge  != 5)
covar_para <- covar %>% filter(challenge == 3, re_challenge  != 2)

# Pheno updates
pheno_enteric <- covar_enteric[,c(1,5)]
pheno_typhoid <- covar_typhoid[,c(1,5)]
pheno_para <- covar_para[,c(1,5)]

# Check sample covar pheno ID info ---------------------------------------------
fam <- fread("merge_rn.fam")
fam <- fam %>% rename(IID = V2)
missing <- anti_join(fam, covar_enteric, by = "IID")
missing$IID_2 <- str_remove_all(missing$IID, "^([0-9]+-)")
#See if data is in old covar files
check <- filter(covar2, GENOTYPING_ID %in% missing$IID_2) # Found 5 samples :)
check <- filter(check, !is.na(Diagnosed))
check <- distinct(check)

check2 <- filter(missing, IID_2 %in% covar2$GENOTYPING_ID)


# Tidy data for updating in covarupdating script
check <- rename(check, IID = GENOTYPING_ID)
check$Vaccine <- ifelse(check$Vaccine == "None",0,
                            ifelse(check$Vaccine == "Ty21a",1,
                                   ifelse(check$Vaccine == "M01ZH09", 2,
                                          ifelse(check$Vaccine == "Vi-PS", 3,
                                                 ifelse(check$Vaccine == "Vi-TT", 4,NA)))))
#Recode strain
check$Challenge <- as.factor(check$Challenge)
levels(check$Challenge)
#C, P = 3, T = 1, TN = 2
levels(check$Challenge) <- c("3", "1")
#sex M=1, F=2
check$Sex <- ifelse(check$Sex == "M", 1, ifelse(check$Sex == "F", 2, NA))
#Re-challenge
check$Re_Challenge <- as.factor(check$Re_Challenge)
#N=1, PP=2, PT=3, TP=4, TT=5
levels(check$Re_Challenge) <- c("1")
check <- clean_names(check)
check <- rename(check, IID_2 = iid)

id <- check2 %>% select(V1, IID, IID_2)

update <- left_join(check, id, by = "IID_2")
update <- update[,c(20,2:18)]
# add to covar_enteric
update$challenge <- as.integer(update$challenge)
update$re_challenge <- as.integer(update$re_challenge)

covar<- bind_rows(covar, update)

# Remake files
# Filter
covar_enteric <- covar %>% filter(re_challenge == 1)
covar_typhoid <- covar %>% filter(challenge != 3, re_challenge  != 5)
covar_para <- covar %>% filter(challenge == 3, re_challenge  != 2)

# Pheno updates
pheno_enteric <- covar_enteric[,c(1,5)]
pheno_typhoid <- covar_typhoid[,c(1,5)]
pheno_para <- covar_para[,c(1,5)]


# recheck
'%!in%' <- function(x,y)!('%in%'(x,y))
fam_miss_covar <- fam[(fam$IID %!in% covar_enteric$IID),]
# Still missing 010, 047, 078, 087, 002
# zeros removed from IDs 

# Just need to rename covar file 
extra <- c("010", "047", "078", "087", "002") #did in excel bc i am tired
write.csv(covar, "covar.csv", row.names = F)
covar <- fread("covar.csv")


# Make keep file
keep_t <- inner_join(covar_typhoid, fam, by = "IID")
keep_t <- select(keep_t, V1, IID)

keep_p <- inner_join(covar_para, fam, by = "IID")
miss <- anti_join(covar_para, fam, by = "IID") #some have patch IDs so need to get old ones
keep_p <- select(keep_p, V1, IID)

write.table(keep_t, file="keep_t.txt",col.names = F, row.names = F, quote = F, sep = '\t')
write.table(keep_p, file="keep_p.txt",col.names = F, row.names = F, quote = F, sep = '\t')

files <- mget(ls())

for (i in 1:length(files)){
  write.table(files[[i]], paste(names(files[i]), ".txt", sep = ""), row.names = F, quote = F, sep = '\t')
}
# move to ~/GWAS_22/gwas_final/meta

covar$study <- as.factor(covar$study)
table(covar$study)
# Samples
#P1 PATCH    T1    T2 TYGER  VAST 
#50    36    38    80    40   102
```
#### Typhoid


```{bash filter typhoid}
plink2 --bfile merge_rn --keep keep_t.txt --make-bed --out typhoid
plink2 --bfile merge_rn --keep keep_p.txt --make-bed --out paratyphoid
```


#### Paratyphoid

#### All enteric







### PCA

-   Use a PCA to account for population structure in data

-   Check ancestry first by plotting PCA and colour coding by ethnicity

-   Can also merge with public population databases

-   Include PCA's as covariates in assocation model

```{bash Run PCA}
plink2 --bfile all_enteric_QC2 --pca
```

```{r Set up PCA in R, eval=TRUE}
pca <- fread("plink2.eigenvec")
library("ggplot2")

ggplot(pca, aes(PC1,PC2)) + geom_point()
eigenval <- fread("plink2.eigenval")

#Colour by ethnicity

covar <- fread("all_covar_num.txt")

pca_covar <- left_join(pca, covar, by = "IID")
#matched 335 rows, 13 rows in x without a match, 11 rows in covar without a match?
missing <- anti_join(pca, covar, by = "IID")
extra_covar <- anti_join(covar, pca, by = "IID")
#Need to add extra 00s for T2 samples in covar file
#now 8 rows missing from covar, probs deleted participants plus the samples we couldn't find covar or pheno info for anywhere
#Double check repeat samples later

pca_covar$Ethnicity <- as.factor(pca_covar$Ethnicity)
levels(pca_covar$Ethnicity)
#merge white british into one factor

pca_covar$Ethnicity <- recode_factor(pca_covar$Ethnicity, "White British" = "White (British)")
pca_covar <- pca_covar %>% mutate_all(na_if,"")

#Plot PCA coloured by ethnicity
pca_covar %>% ggplot(aes(PC1,PC2, color=Ethnicity)) + geom_point()

#Calculate Percentage variance explained

pve <- data.frame(PC = 1:10, pve = round(eigenval/sum(eigenval)*100,2))

colnames(pve) <- c("PC", "pve")
pve$PC <- as.factor(pve$PC)

p <- pve %>% ggplot(aes(x=PC,y=pve, fill = PC)) + geom_col()
p + labs(x = "Principle Component", y = "Percentage Varience Explained")
#Make new covar file to use
#write.table(pca_covar, file = "pca_covar.txt", row.names = F, quote = F, sep = "\t")

```

### Conclusions

-   Include first 5 PC's as covariates in association analysis
